---
title: "hw3_AashwinLamsal"
author: "Aashwin Lamsal"
date: "10/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Homework 3: NYC Flights Data
This data frame contains all 336,776 flights that departed from New York City in 2013. The data comes from the US Bureau of Transportation Statistics, and is documented in ?flights.

1) Find all flights that:
-Had an arrival delay of two or more hours
-Flew to Houston (IAH or HOU)
-Departed in summer (July, August, and September)
-Arrived more than two hours late, but didn’t leave late
-Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
library(tidyverse)
data(flights,package="nycflights13")
head(flights)

#flights that had an arrival delay of two or more hours
arrDelay2Hours <- filter(flights, arr_delay>=120)#since delay times are given in minutes
head(arrDelay2Hours, 10)

#flights that flew to Houston airports: IAH or HOU
houstonFlights <- filter(flights, dest == "IAH"|dest == "HOU")
head(houstonFlights, 10)

#flights that departed in the summer monthts: July, August, or September
summerFlights <- filter(flights, month == 7 | month == 8 | month == 9)
head(summerFlights, 10)

#flights that arrived more than 2 hours late, but did NOT leave late
arriveLateLeaveOnTime <- filter(flights, arr_delay>120 & dep_delay<=0)
head(arriveLateLeaveOnTime,10)

#flights that were delayed by AT LEAST an hour, but made up OVER 30 minutes in flight
makeUp30 <- filter(flights, dep_delay >= 60 & (dep_delay - arr_delay > 30)) # if departure delay is 10 minutes late, we expect arrival delay to be 10 minutes late as well
#but if arrival delay is -20 minutes, that means the flight caught up by 30 minutes in the air (10- (-20) = 30)
head(makeUp30, 10)
```
2) From lecture, our definition of cancelled flights (is.na(dep_delay) | is.na(arr_delay) ). This is not optimal. Why? Explain which is the most important column.
    -  The current definition of cancelled flights is not optimal because if there is
       an instance of NA values showing up in dep_delay, then there isn't a feasible 
       way for their to be a non-NA value in arr_delay, so having the conditional
       statement isn't optimal. This is why the dep_delay column is the most 
       important column between the two, since the dep_delay affects the arr_delay 
       column the most, ideally the arr_delay should be equally delayed as the 
       dep_delay (if your flight was delayed by 10 minutes, you would expect to arrive 10 minutes late).
       
       
3) What time of day should you fly if you want to avoid delays as much as possible?
```{r}
library(tidyverse)
library(ggplot2)
data(flights,package="nycflights13")
zeroDelays <- filter(flights, (dep_delay <= 0 & arr_delay <= 0))
timesWithNoDelay <- select(zeroDelays,hour, minute, time_hour, dep_delay)
head(timesWithNoDelay, 15)

ggplot(timesWithNoDelay, aes(hour, group = time_hour)) + geom_histogram() + labs(x = "Hour of Day", y = "Frequency of Flights", title = "Distribution of Times for Flights With No Delays")


```




After the conditional statment used filtered the dataframe by about half of its original size. Sorting the dataframe by hour showed that most of the flight times where delays were avoided took place between 5 and 10 AM. There were times that had avoidable delays later on in the day. but they weren't as frequent (judging infrequency by time_hour to see the dates of the flights) as the early morning flights.

4) There are 4 other relational datasets in the nycflights13 package,
airlines gives information on the abbreviated airport codes,
airports gives info on each airport, identified by the faa airport code,
planes gives info on each plane, identified by its tailnum,
weather gives info on each NYC airport for each hour.

Loading the datasets...
```{r}
data(airlines,package='nycflights13')
data(airports,package='nycflights13')
data(planes,package='nycflights13')
data(weather,package='nycflights13')

```
After viewing the diagram, write out how the dataset flights connects to these 4 other datasets (there may be more than one variable that connects them). I’ll do the first one for you:

- flights connects to planes through a single variable, tailnum.
Your turn:
- flights connects to airlines through a single variable, carrier.
- flights connects to airports through two variables, origin and dest.
- flights connects to weather through five variables, year, month, day, hour, and origin.

5) The relationship between weather and airports is the single variable origin that they share. On the diagram, an arrow should connect origin in the weather table to faa/origin in the airports table.

6) The following code joins data frames flights and weather together. Based on flight_weather, what weather conditions make it more likely to see a delay?
```{r}
library(ggplot2)
library(tidyverse)
data(airlines,package='nycflights13')
data(airports,package='nycflights13')
data(planes,package='nycflights13')
data(weather,package='nycflights13')


flight_weather = flights %>%
inner_join(weather, by = c(
  "origin" = "origin",
  "year" = "year",
  "month" = "month",
  "day" = "day",
  "hour" = "hour"
 ))

head(flight_weather)

weatherDelays <- filter(flight_weather, (dep_delay > 60 | arr_delay > 60))

head(weatherDelays)

ggplot(weatherDelays, aes(precip)) + geom_histogram(bins=20) + labs(x = "Precipitation (in)", y="No. of Flights", title = "Distribution of Flights by Amount of Precipitation") #distribution of the number of flights versus precipitation amounts

ggplot(weatherDelays, aes(visib)) + geom_histogram(bins=20) + labs(x = "Visibility (mi)", y="No. of Flights", title = "Distribution of Flights by Visible Distance") #distribution of the number of flights versus visiblity

ggplot(weatherDelays, aes(wind_speed)) + geom_histogram(bins=20) + labs(x = "Wind Speed (mph)", y="No. of Flights", title = "Distribution of Flights by Wind Speed") #distribution of the number of flights versus precipitation amounts




```

Based on what was seen in the distributions above, the weather conditions that affect weather delays the most seem to be visibility, precipitation, and wind speed. Visibility must be around 10 miles, otherwise the plane will not take off. It's a similar situation for precipitation, as most flights will not take off if there is any storms/rainfall happening at the airport. Wind speed however is slightly different, mainly because most flights take off when the wind is between 10 and 15 miles per hour. Flights will continue to take off with winds upwards of 25 miles per hour, though the number of flights do dwindle, and after winds reach speeds of 30 miles per hour, there are no planes taking off.

7) If we had used left_join in the previous problem, would that change our results? Investigate. Does using an inner_join or left_join make more sense here?
```{r}
library(ggplot2)
library(tidyverse)
data(airlines,package='nycflights13')
data(airports,package='nycflights13')
data(planes,package='nycflights13')
data(weather,package='nycflights13')


flightLeft = flights %>%
left_join(weather, by = c(
  "origin" = "origin",
  "year" = "year",
  "month" = "month",
  "day" = "day",
  "hour" = "hour"
 ))

head(flightLeft)

weatherLeftDelays <- filter(flightLeft, (dep_delay > 60 | arr_delay > 60))

head(weatherLeftDelays)

ggplot(weatherLeftDelays, aes(precip)) + geom_histogram(bins=20) + labs(x = "Precipitation (in)", y="No. of Flights", title = "Distribution of Flights by Amount of Precipitation") #distribution of the number of flights versus precipitation amounts

ggplot(weatherLeftDelays, aes(visib)) + geom_histogram(bins=20) + labs(x = "Visibility (mi)", y="No. of Flights", title = "Distribution of Flights by Visible Distance") #distribution of the number of flights versus visiblity

ggplot(weatherLeftDelays, aes(wind_speed)) + geom_histogram(bins=20) + labs(x = "Wind Speed (mph)", y="No. of Flights", title = "Distribution of Flights by Wind Speed") #distribution of the number of flights versus precipitation amounts



```


In a left-join, all data in the left table (in our case, this would be flights) are kept, with nonmatches between the flights and weather tables being filled in by NA values. However, it seems that it doesn't matter whether we use an inner-join or a left-join, since both combined tables share all of the same attributes/variables. As far as which join makes more sense goes, inner-joins only keep matches between both tables, while all outer-joins, including left-joins, will keep all data in one of the two tables, so I think that when combining tables, you would want to look for matches between the two tables, so I would recommend inner-join over left-join.

8) Is there a relationship between the age of a plane and its departure delays? Information on age of a plane can be found in planes under variable year. Information on departure delays can be found in flights under variable dep_delay.

```{r}
planes <- mutate(planes, "age" = 2019 - year)

flightsPlanesJoin <-
  planes %>%
  inner_join(flights, by = c("tailnum" = "tailnum"))

head(flightsPlanesJoin)
ggplot(flightsPlanesJoin, aes(x = age, y = dep_delay)) + geom_point() + labs(x="Plane Age (yrs)", y = "Departure Delay (min)", title = "Delays versus Plane Age")

```
Judging from the scatter plot made above, there seems to be no correlation between the age of the plane and the length of a delay, in fact older planes seem to have shorter delays than planes age 0-25.

9) What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?

  -Anti-joins between two tables are primarily used to look for join mismatches, since anti_join(x,y) drops all observations in x that have a match in y.

```{r}
flightsAirports <- inner_join(flights, airports, by = c("dest"="faa"))
flightsAnti <- anti_join(flights, airports, by = c("dest" = "faa"))
head(flightsAirports)
head(flightsAnti)
head(flights)

```
Anti-joining between flights and airports didn't drop any values from the flight table at all, which tells me that airports and flights don't have any corresponding matching values that can be dropped in the first place, aside from dest and faa since their its the primary key between the two tables.

```{r}
flightsAirports2 <- inner_join(airports, flights, by = c("faa"="dest"))
flightsAnti2 <- anti_join(airports, flights, by = c("faa" = "dest"))

head(flightsAirports2)
head(flightsAnti2)
head(flights)

```
Anti-joining between airports and flights, this time switching the order of the join so we also have to switch the order of the column names that correspond to the correct tables, shows me only columns found in the airports table. This further shows that airports and flights don't have any other corresponding columns/values, aside from the primary key that they share (faa and dest)


















