---
title: "hw3_AashwinLamsal"
author: "Aashwin Lamsal"
date: "10/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Homework 3: NYC Flights Data
This data frame contains all 336,776 flights that departed from New York City in 2013. The data comes from the US Bureau of Transportation Statistics, and is documented in ?flights.

1) Find all flights that:
-Had an arrival delay of two or more hours
-Flew to Houston (IAH or HOU)
-Departed in summer (July, August, and September)
-Arrived more than two hours late, but didn’t leave late
-Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
library(tidyverse)
data(flights,package="nycflights13")
head(flights)

#flights that had an arrival delay of two or more hours
arrDelay2Hours <- filter(flights, arr_delay>=120)#since delay times are given in minutes
head(arrDelay2Hours, 10)

#flights that flew to Houston airports: IAH or HOU
houstonFlights <- filter(flights, dest == "IAH"|dest == "HOU")
head(houstonFlights, 10)

#flights that departed in the summer monthts: July, August, or September
summerFlights <- filter(flights, month == 7 | month == 8 | month == 9)
head(summerFlights, 10)

#flights that arrived more than 2 hours late, but did NOT leave late
arriveLateLeaveOnTime <- filter(flights, arr_delay>120 & dep_delay<=0)
head(arriveLateLeaveOnTime,10)

#flights that were delayed by AT LEAST an hour, but made up OVER 30 minutes in flight
makeUp30 <- filter(flights, dep_delay >= 60 & (dep_delay - arr_delay > 30)) # if departure delay is 10 minutes late, we expect arrival delay to be 10 minutes late as well
#but if arrival delay is -20 minutes, that means the flight caught up by 30 minutes in the air (10- (-20) = 30)
head(makeUp30, 10)


```

